---
title: "Lab 5: Resampling Methods"
author: "Rich Pauloo"
date: "6/24/2018"
output: md_document
---

# Lab 5: Resampling Methods

*6/24/2018*  

***  

```{r setup, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, error = FALSE)
```

## 5.3.1 The Validation Set Approach

We will use the built-in `Auto` dataset in ISLR. We take half of the 392 observations as training data and fit a linear model.

```{r}
library(ISLR)            # for datasets
set.seed(1)              # for reproducibility
train <- sample(392,196) # random subset of 196 obs. from original 392 obs

# fit the linear model
lm_fit <- lm(mpg ~ horsepower, 
             data = Auto, 
             subset = train)
```

We now use the `predict()` function to estimate the response for all 392 observations, and we use the `mean()` function to calculate the MSE of the 196 observations in the validation set. Note that the `-train` index below selects only the observations that are not in the training set.
```{r}
# predict the auto data and subset for the test data
test_pred <- predict(lm_fit, newdata = Auto)[-train]

# calculate the mean squared error
mean((test_pred - Auto$mpg[-train])^2)

```

Therefore, the estimated test MSE for the linear regression fit is 26.14. We can use the `poly()` function to estimate the test error for the polynomial and cubic regressions.

```{r}
# polynomial second degree fit 
lm_fit2 <- lm(mpg ~ poly(horsepower, 2),
              data = Auto,
              subset = train)

# MSE
mean((Auto$mpg - predict(lm_fit2, Auto))[-train]^2)

# polynomial third degree fit 
lm_fit3 <- lm(mpg ~ poly(horsepower, 3),
              data = Auto,
              subset = train)

# MSE
mean((Auto$mpg - predict(lm_fit3, Auto))[-train]^2)
```

These error rates are 19.82 and 19.78, respectively. If we choose a different training set instead, then we will obtain somewhat different errors on the validation set.

```{r}
set.seed(2)              # reproducibility
train <- sample(392,196) # new training sample

# re-fit model
lm_fit <- lm(mpg ~ horsepower,
             data = Auto,
             subset = train)

# MSE
mean((Auto$mpg - predict(lm_fit, Auto))[-train]^2)

# polynomial second degree fit 
lm_fit2 <- lm(mpg ~ poly(horsepower, 2),
              data = Auto,
              subset = train)

# MSE
mean((Auto$mpg - predict(lm_fit2, Auto))[-train]^2)

# polynomial third degree fit 
lm_fit3 <- lm(mpg ~ poly(horsepower, 3),
              data = Auto,
              subset = train)

# MSE
mean((Auto$mpg - predict(lm_fit3, Auto))[-train]^2)
```

Using this split of the observations into a training set and a validation set, we find that the validation set error rates for the models with linear, quadratic, and cubic terms are **23.30**, **18.90**, and **19.26**, respectively.  

These results are consistent with our previous findings: a model that predicts `mpg` using a quadratic function of `horsepower` performs better than a model that involves only a linear function of `horsepower`, and there is little evidence in favor of a model that uses a cubic function of `horsepower`.  

***  

## 5.3.2 Leave-One-Out Cross-Validation

`cv.glm()`, part of the `boot` library, computes the LOOCV estimate for `glm()` models. `glm()` without a `family=""` argument defaults to a lienar model. Therefore the following two models are equivalent. We use `glm()` in this section because `cv.glm()` works with glm objects. 
```{r}
m <- glm(mpg ~ horsepower, data = Auto) # fit a linear model with `glm`
coef(m) # model coefficients

m2 <- lm(mpg ~ horsepower, data = Auto) # fit a linear model with `lm`
coef(m2) # model coefficients
```

```{r}
library(boot) # for LOOCV : `cv.glm()`

glm_fit <- glm(mpg ~ horsepower, # fit model
               data = Auto)

cv_err = cv.glm(Auto, glm_fit)
cv_err$delta
```

The `delta` element gies us cross validation results: the LOOCV estimate, which is defined as the average test MSE estimates.  

$CV_{n} = \frac{1}{n}\sum_{i = 1}^{n}MSE_{i}$   

We can repeat this procedure for increasingly complex polynomial fits. To automate the process, we use the `for()` function to initiate a for loop which iteratively fits polynomial regressions for polynomials of order i = 1 to i = 5, computes the associated cross-validation error, and stores it in the ith element of the vector cv.error. We begin by initializing the vector. 

```{r}
cv_error = rep(0, 5)

# fit glm with increasing degree of polynomial
for (i in 1:5){
  glm_fit = glm(mpg ~ poly(horsepower, i), 
                data = Auto)
  cv_error[i] = cv.glm(Auto, glm_fit)$delta[1]
}

# plot
plot(1:5, cv_error, 
     main = "LOOCV MSE with increasing polynomial fits",
     xlab = "Degree Polynomial",
     ylab = "LOOCV Error")
```

As in Figure 5.4, we see a sharp drop in the estimated test MSE between the linear and quadratic fits, but then no clear improvement from using higher-order polynomials.  

***  

## 5.3.3 k-Fold Cross Validation

The `cv.glm()` function can also be used to implement k-fold CV. Below we use *k = 10*, a common choice for *k*, on the `Auto` data set. We once again set a random seed and initialize a vector in which we will store the CV errors corresponding to the polynomial fits of orders one to ten.

```{r}
set.seed(17)

cv.error.10 = rep(0,10) # vector to fill

for (i in 1:10){
  # fit the model with increasing polynomials
  glm.fit = glm(mpg ~ poly(horsepower, i), data = Auto)
  # store the raw cross-validation estimate of prediciton error
  cv.error.10[i] = cv.glm(Auto, glm.fit, K = 10)$delta[1] 
}

cv.error.10
```
We still see little evidence that using cubic or higher-order polynomial terms leads to lower test error than simply using a quadratic fit.  

***  

## 5.3.4 The Bootstrap

### Estimating the Accuracy of a Statistic of Interest


One of the great advantages of the bootstrap approach is that it can be applied in almost all situations. No complicated mathematical calculations are required. Performing a bootstrap analysis in R entails only two steps. First, we must create a function that computes the statistic of interest. Second, we use the the `boot()` function, which is part of the boot library, to perform the bootstrap by repeatedly sampling observations from the data set with replacement.  

The Portfolio data set in the ISLR package is described in Section 5.2. To illustrate the use of the bootstrap on this data, we must first create a function, `alpha.fn()`, which takes as input the (X, Y) data as well as a vector indicating which observations should be used to estimate $\alpha$. The function then outputs the estimate for α based on the selected observations.  

```{r}
alpha.fn = function(data, index){ 
  X = data$X[index]
  Y = data$Y[index]
  return((var(Y) - cov(X,Y)) / (var(X) + var(Y) - 2*cov(X,Y)))
}
```

This function returns, or outputs, an estimate for α based on applying (5.7) to the observations indexed by the argument index. For instance, the following command tells R to estimate α using all 100 observations.

```{r}
alpha.fn(Portfolio,1:100)
```
The next command uses the `sample()` function to randomly select 100 observations from the range 1 to 100, with replacement. This is equivalent to constructing a new bootstrap data set and recomputing $\hat{\alpha}$ based on the new data set.

```{r}
set.seed(1)
alpha.fn(Portfolio, sample(100, 100, replace = T))
```

We can implement a bootstrap analysis by performing this command many times, recording all of the corresponding estimates for $\alpha$, and computing the resulting standard deviation. However, the `boot()` function automates this approach. Below we produce R = 1,000 bootstrap estimates for $\alpha$.

```{r}
boot(Portfolio, alpha.fn, R = 1000)
```
The final output shows that using the original data, $\hat{\alpha}$ = 0.5758, and that the bootstrap estimate for $SE(\hat{\alpha})$ is 0.0886.

***  

### Estimating the Accuracy of a Linear Regression Model

The bootstrap approach can be used to assess the variability of the coefficient estimates and predictions from a statistical learning method. Here we use the bootstrap approach in order to assess the variability of the estimates for $\beta_0$ and $\beta_1$, the intercept and slope terms for the linear regression model that uses horsepower to predict mpg in the Auto data set. We will compare the estimates obtained using the bootstrap to those obtained using the formulas for \\(SE(\hat{\beta_0})\\) and $SE(\hat{\beta_1})$ described in Section 3.1.2.  

We first create a simple function, `boot.fn()`, which takes in the Auto data set as well as a set of indices for the observations, and returns the intercept and slope estimates for the linear regression model. We then apply this function to the full set of 392 observations in order to compute the estimates of $\beta_0$ and $\beta_1$ on the entire data set using the usual linear regression coefficient estimate formulas from Chapter 3. 

```{r}
boot.fn = function(data,index){
  return(coef(lm(mpg ~ horsepower, data = data, subset = index)))
}

boot.fn(Auto, 1:392)
```

The `boot.fn()` function can also be used in order to create bootstrap estimates for the intercept and slope terms by randomly sampling from among the observations with replacement. Here we give two examples.

```{r}
set.seed(1)
boot.fn(Auto, sample(392, 392, replace = T)) 

boot.fn(Auto, sample(392, 392, replace = T))
```

Next, we use the `boot()` function to compute the standard errors of 1,000 bootstrap estimates for the intercept and slope terms.
```{r}
boot(Auto, boot.fn, 1000)
```

This indicates that the bootstrap estimate for $SE(\hat{\beta_0})$ is 0.86, and that the bootstrap estimate for $SE(\hat{\beta_1})$ is 0.0074. As discussed in Section 3.1.2, standard formulas can be used to compute the standard errors for the regression coefficients in a linear model. These can be obtained using the `summary()` function.

```{r}
summary(lm(mpg ~ horsepower, data = Auto))$coef
```

The standard error estimates for $\hat{\beta_0}$ and $\hat{\beta_1}$ obtained using the formulas from Section 3.1.2 are 0.717 for the intercept and 0.0064 for the slope. Interestingly, these are somewhat different from the estimates obtained using the bootstrap. Does this indicate a problem with the bootstrap? In fact, it suggests the opposite. Recall that the standard formulas given in Equation 3.8 on page 66 rely on certain assumptions. For example, they depend on the unknown parameter $\sigma^2$, the noise variance. We then estimate $\sigma^2$ using the RSS. Now although the formula for the standard errors do not rely on the linear model being correct, the estimate for $\sigma^2$ does. We see in Figure 3.8 on page 91 that there is a non-linear relationship in the data, and so the residuals from a linear fit will be inflated, and so will $\sigma^2$,. Secondly, the standard formulas assume (somewhat unrealistically) that the xi are fixed, and all the variability comes from the variation in the errors $\epsilon_i$. The bootstrap approach does not rely on any of these assumptions, and so it is likely giving a more accurate estimate of the standard errors of $\hat{\beta_0}$ and $\hat{\beta_1}$ than is the `summary()` function.  

Below we compute the bootstrap standard error estimates and the standard linear regression estimates that result from fitting the quadratic model to the data. Since this model provides a good fit to the data (Figure 3.8), there is now a better correspondence between the bootstrap estimates and the standard estimates of $SE(\hat{\beta_0})$, $SE(\hat{\beta_1})$ and $SE(\hat{\beta_2})$.  
```{r}
boot.fn = function(data, index){
  # rewrite function as quadratic polynomial
  coefficients(lm(mpg ~ horsepower + I(horsepower^2), data = data, subset = index))
}

# obtain bootstrap estaimtes of standard errors of coefficients
set.seed(1)
boot(Auto, boot.fn, 1000)

# compare to emperical calculations of SE
summary(lm(mpg ~ horsepower + I(horsepower^2), data = Auto))$coef
```

